---
title: Expression in tumor and nontumor tissues
subtitle: 5mhC expression in bladder cancer
output: 
  html_document: 
    keep_md: yes
---

```{r Global, echo = FALSE, warning = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(plyr)
library(dplyr)
library(tidyr)
source("gplots.R")
```

```{r Data}
# Loading the dataset, libraries and sources for analysis
data <- read.delim("DatasetBladder.csv")
```

In this section we will compare 5hmC H-scores as they were estimated in paired (i.e, form the sama surgical specimen) normal (i.e., histologically-normal urothelium) and tumor (both invasive and noninvasive) tissues. For this we will use the full dataset. Comparisons were made using the Wilcoxon signed rank test.

```{r}
# Dataframe for mean expression
mean_5hmc <- data %>%
  select(t.5hmc.m, n.5hmc.m) %>%
  gather(Tissue, H, t.5hmc.m:n.5hmc.m) %>%
  mutate(Tissue = ifelse(Tissue == "t.5hmc.m", c("Tumor"), c("Normal")))
# Mean 5hmC expression in normal and tumor tissues
var <- mean_5hmc$H
group <- mean_5hmc$Tissue
gbox(var, group, mean_5hmc, pair = TRUE) +
  ggtitle("Mean 5hmC expression in normal and \ntumor tissues") +
  ylab("Mean 5hmC score")
# Dataframe for median expression
median_5hmc <- data %>%
  select(contains("5hmc.med")) %>%
  gather(Tissue, H, t.5hmc.med:n.5hmc.med) %>%
  mutate(Tissue = ifelse(Tissue == "t.5hmc.med", c("Tumor"), c("Normal")))
# Median 5hmC expression in normal and tumor tissues
var <- median_5hmc$H
group <- median_5hmc$Tissue
gbox(var, group, median_5hmc, pair = TRUE) +
  ggtitle("Median 5hmC expression in normal and \ntumor tissues") +
  ylab("Median 5hmC score")
```

***

## 5hmC expression in invasive and noninvasive tumors
In addition, we will compare mean and median 5hmC scores between noninvasive and invasive tumors. Mean and median H-scores will be obtained from `t.5hmc.m` and `t.5hmc.med` (respectively), using `spot.dx` to identify which cases corresponded to noninvasive or invasive tumors. Regarding the latter, in some cases there was a discrepancy between the diagnosis given in the pathology report (as specified in `histo.dx`) and what was found at the TMA spot (specified in `spot.dx`). The following plot shows what was found at the TMA spots.

```{r spotDx}
var <- data$spot.dx
gbar(var, data)
```

Thus, for the next comparisons and plots we will use only the `Invasive` and `Noninvasive` cases. Comparisons were made using the Mann-Whitney U test.

```{r}
# Dataset
data_tumor <- data %>% filter(spot.dx != "Normal") %>% droplevels()
# Mean H-scores
var <- data_tumor$t.5hmc.m
group <- data_tumor$spot.dx
gbox(var, group, data_tumor) +
  ggtitle("Mean 5hmC expression in invasive and \noninvasive tumors") +
  ylab("Mean 5hmC score")
# Median H-scores
var <- data_tumor$t.5hmc.med
group <- data_tumor$spot.dx
gbox(var, group, data_tumor) +
  ggtitle("Median 5hmC expression in invasive and \noninvasive tumors") +
  ylab("Median 5hmC score")
```

***
