---
title: Expression and Outcome
subtitle: 5mhC expression in bladder cancer
output: 
  html_document: 
    keep_md: yes
---

```{r Global, echo = FALSE, warning = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(plyr)
library(dplyr)
library(tidyr)
source("gplots.R")
```

```{r Data}
# Loading the dataset, libraries and sources for analysis
data <- read.delim("DatasetBladder.csv")
# Lumping pT levels at cystectomy considering muscle invasion
data <- data %>%
  mutate(mid = ifelse(pt.cystectomy == "Tis-Ta" | pt.cystectomy == "T1", "No", "Yes"))
data$mid <- factor(data$mid)
```

For this section we will use 3 different approaches for evaluating the association of 5hmC and outcome: 1) using 5hmC H-scores; 2) categorizing 5hmC expression in low/high levels using the median as the cutoff point; 3) categorizing 5hmC expression in low/high levels using the Q3 as the cutoff point.

Only invasive cases will be consider from now on, as observed at the TMA spot (i.e., `spot.dx == Invasive`). For age (numeric variable) we will use the median to define younger and older age groups.

```{r Invasive}
data_tumor <- data %>%
  filter(spot.dx == "Invasive") %>%
  mutate(age_group = ifelse(age > median(age), "Older", "Younger")) %>%
  droplevels()
data_tumor$age_group <- factor(data_tumor$age_group, ordered = TRUE)
```

The number of invasive cases is N = `r nrow(data_tumor)`. Also, only median H-scores will be consider for analysis.

## Approach #1
In this approach we will compare 5hmC `H-scores` against clinicopathologic features and outcome. Comparisons will be made using the Mann-Whitney U test. 

```{r Approach1}
# For all plots in this section
data <- data_tumor
var <- data$t.5hmc.med
# 5hmC and patient's age group
group <- data$age_group
gbox(var, group) + 
  ggtitle("5hmC expression by group age") +
  ylab("Median 5hmC score")
# 5hmC and patient's sex
group <- data$sex
gbox(var, group) + 
  ggtitle("5hmC expression by patient's sex") +
  ylab("Median 5hmC score")
# 5hmC and patient's race
group <- data$race
gbox(var, group) + 
  ggtitle("5hmC expression by patient's race") +
  ylab("Median 5hmC score")
# 5hmC and pT stage at cystectomy
group <- data$mid
gbox(var, group) +
  ggtitle("5hmC expression by presence of \nmuscle-invasive (>pT1) tumor") +
  ylab("Median 5hmC score")
# 5hmC and lymphovascular invasion
group <- data$lvi
gbox(var, group) + 
  ggtitle("5hmC expression by presence of \nlymphovascular invasion") +
  ylab("Median 5hmC score")
# 5hmC and lymph nodes metastases
group <- data$ln.met
gbox(var, group) + 
  ggtitle("5hmC expression by presence of \nlymph node metastases") +
  ylab("Median 5hmC score")
# 5hmC and early complications
group <- data$comp.early
gbox(var, group) + 
  ggtitle("5hmC expression by presence of \nearly complications") +
  ylab("Median 5hmC score")
# 5hmC and late complications
group <- data$comp.late
gbox(var, group) + 
  ggtitle("5hmC expression by presence of \nlate complications") +
  ylab("Median 5hmC score")
# 5hmC and metastatic disease after cystectomy
d <- data_tumor %>% filter(!is.na(metastasis))
var <- d$t.5hmc.med
group <- d$metastasis
gbox(var, group, d) + 
  ggtitle("5hmC expression by presence of \nmetastatic disease after cystectomy") +
  ylab("Median 5hmC score")
# 5hmC and overall mortality
var <- data$t.5hmc.med
group <- data$death
gbox(var, group) + 
  ggtitle("5hmC expression by overall mortality") +
  ylab("Median 5hmC score")
# 5hmC and cancer-related mortality
d <- data_tumor %>% filter(!is.na(dod))
var <- d$t.5hmc.med
group <- d$dod
gbox(var, group, d) + 
  ggtitle("5hmC expression by cancer-related mortality") +
  ylab("Median 5hmC score")
# 5hmC and tumor progression
var <- data$t.5hmc.med
group <- data$progression
gbox(var, group) + 
  ggtitle("5hmC expression by presence of \ntumor progression") +
  ylab("Median 5hmC score")
```

***

## Approach # 2
In this approach we will compare 5hmC levels (categorizing 5hmC expression in low/high levels using the `median` as the cutoff point) against clinicopathologic features and outcome. Comparisons will be made using the Fisher's exact test.

```{r Approach2}
# Splitting in 2 levels based on median score
data_tumor <- data_tumor %>%
  mutate(p50 = ifelse(t.5hmc.med > median(t.5hmc.med, na.rm = TRUE), "High", "Low")) %>%
  droplevels()
# For all plots in this section
var2 <- data_tumor$p50
# 5hmC levels and patient's age group
var1 <- data_tumor$age_group
gbar2(var1, var2) + ggtitle("5hmC levels and patient's age group")
# 5hmC levels and patient's sex
var1 <- data_tumor$sex
gbar2(var1, var2) + ggtitle("5hmC levels and patient's sex")
# 5hmC levels and patient's race
var1 <- data_tumor$race
gbar2(var1, var2) + ggtitle("5hmC levels and patient's race")
# 5hmC and pT stage at cystectomy
var1 <- data_tumor$mid
gbar2(var1, var2) + ggtitle("5hmC levels by presence of \nmuscle-invasive (>pT1) tumor")
# 5hmC and lymphovascular invasion
var1 <- data_tumor$lvi
gbar2(var1, var2) + ggtitle("5hmC levels by presence of \nlymphovascular invasion")
# 5hmC and lymph nodes metastases
var1 <- data_tumor$ln.met
gbar2(var1, var2) + ggtitle("5hmC levels by presence of \nlymph node metastases")
# 5hmC and early complications
var1 <- data_tumor$comp.early
gbar2(var1, var2) + ggtitle("5hmC levels by presence of \nearly complications")
# 5hmC and late complications
var1 <- data_tumor$comp.late
gbar2(var1, var2) + ggtitle("5hmC levels by presence of \nlate complications")
# 5hmC and metastatic disease after cystectomy
var1 <- data_tumor$metastasis
gbar2(var1, var2) + ggtitle("5hmC levels by presence of \nmetastatic disease after cystectomy")
# 5hmC and overall mortality
var1 <- data_tumor$death
gbar2(var1, var2) + ggtitle("5hmC levels by overall mortality")
# 5hmC and cancer-related mortality
var1 <- data_tumor$dod
gbar2(var1, var2) + ggtitle("5hmC levels by cancer-related mortality")
# 5hmC and tumor progression
var1 <- data_tumor$progression
gbar2(var1, var2) + ggtitle("5hmC levels by presence of \ntumor progression")
```

***

## Approach # 3
In this approach we will compare 5hmC levels (categorizing 5hmC expression in low/high levels using the `upper tertile` as the cutoff point) against clinicopathologic features and outcome. Comparisons will be made using the Fisher's exact test.

```{r Approach3}
# Splitting in 2 levels based on median score
data_tumor <- data_tumor %>%
  mutate(q3 = ifelse(t.5hmc.med > quantile(t.5hmc.med, .66, na.rm = TRUE), "High", "Low")) %>%
  droplevels()
# For all plots in this section
var2 <- data_tumor$q3
# 5hmC levels and patient's age group
var1 <- data_tumor$age_group
gbar2(var1, var2) + ggtitle("5hmC levels and patient's age group")
# 5hmC levels and patient's sex
var1 <- data_tumor$sex
gbar2(var1, var2) + ggtitle("5hmC levels and patient's sex")
# 5hmC levels and patient's race
var1 <- data_tumor$race
gbar2(var1, var2) + ggtitle("5hmC levels and patient's race")
# 5hmC and pT stage at cystectomy
var1 <- data_tumor$mid
gbar2(var1, var2) + ggtitle("5hmC levels by presence of \nmuscle-invasive (>pT1) tumor")
# 5hmC and lymphovascular invasion
var1 <- data_tumor$lvi
gbar2(var1, var2) + ggtitle("5hmC levels by presence of \nlymphovascular invasion")
# 5hmC and lymph nodes metastases
var1 <- data_tumor$ln.met
gbar2(var1, var2) + ggtitle("5hmC levels by presence of \nlymph node metastases")
# 5hmC and early complications
var1 <- data_tumor$comp.early
gbar2(var1, var2) + ggtitle("5hmC levels by presence of \nearly complications")
# 5hmC and late complications
var1 <- data_tumor$comp.late
gbar2(var1, var2) + ggtitle("5hmC levels by presence of \nlate complications")
# 5hmC and metastatic disease after cystectomy
var1 <- data_tumor$metastasis
gbar2(var1, var2) + ggtitle("5hmC levels by presence of \nmetastatic disease after cystectomy")
# 5hmC and overall mortality
var1 <- data_tumor$death
gbar2(var1, var2) + ggtitle("5hmC levels by overall mortality")
# 5hmC and cancer-related mortality
var1 <- data_tumor$dod
gbar2(var1, var2) + ggtitle("5hmC levels by cancer-related mortality")
# 5hmC and tumor progression
var1 <- data_tumor$progression
gbar2(var1, var2) + ggtitle("5hmC levels by presence of \ntumor progression")
```

***
