---
title: Expression and Survival
subtitle: 5mhC expression in bladder cancer
output: 
  html_document: 
    keep_md: yes
---

```{r Global, echo = FALSE, warning = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(plyr)
library(dplyr)
library(tidyr)
source("gplots.R")
```

```{r Data}
# Loading the dataset, libraries and sources for analysis
data <- read.delim("DatasetBladder.csv")
# Lumping pT levels at cystectomy considering muscle invasion
data <- data %>%
  mutate(mid = ifelse(pt.cystectomy == "Tis-Ta" | pt.cystectomy == "T1", "No", "Yes"))
data$mid <- factor(data$mid)
```

In this section we will plot survival curves for overall mortality and cancer-related mortality according to clinicopathologic features and 5hmC expression, categorizing 5hmC expression in low/high levels using as cutoff points: 1) the median; 2) the upper tertile. Only invasive cases will be consider, as observed at the TMA spot (i.e., `spot.dx == Invasive`). For age (numeric variable) we will use the median to define younger and older age groups.

```{r Invasive}
data_tumor <- data %>%
  filter(spot.dx == "Invasive") %>%
  mutate(age_group = ifelse(age > median(age), "Older", "Younger")) %>%
  droplevels()
data_tumor$age_group <- factor(data_tumor$age_group, ordered = TRUE)
```

We will use `follow-up from cystectomy` for the time-to-event analysis.

```{r FU}
# Establishing follow-up
FU <- data_tumor$fu.cystectomy
```

***

## Clinicopathologic features and survival
In this section we will plot `Kaplan-Meier curves` for overall mortality and cancer-related mortality considering clinicopathologic features. We will compare survival curves using the `log-rank test`.

### Overall mortality
```{r Clinicopath_Death}
# Establishing outcome
outcome <- data_tumor$death
# Patient's age
status <- data_tumor$age_group
gsurv(FU, outcome, status) + ggtitle("Patient's age and overall mortality")
# Patient's sex
status <- data_tumor$sex
gsurv(FU, outcome, status) + ggtitle("Patient's sex and overall mortality")
# Patient's race
status <- data_tumor$race
gsurv(FU, outcome, status) + ggtitle("Patient's race and overall mortality") + theme(legend.position = c(.85, .9))
# pT stage at cystectomy
status <- data_tumor$mid
gsurv(FU, outcome, status) + ggtitle("Presence of muscle-invasive (>pT1) tumor and \noverall mortality")
# Lymphovascular invasion
status <- data_tumor$lvi
gsurv(FU, outcome, status) + ggtitle("Presence of lymphovascular invasion and \noverall mortality")
# Lymph nodes metastases
status <- data_tumor$ln.met
gsurv(FU, outcome, status) + ggtitle("Presence of lymph node metastases and \noverall mortality")
# Early complications
status <- data_tumor$comp.early
gsurv(FU, outcome, status) + ggtitle("Presence of early complications and \noverall mortality")
# Late complications
status <- data_tumor$comp.late
gsurv(FU, outcome, status) + ggtitle("Presence of late complications and \noverall mortality")
# Metastatic disease after cystectomy
status <- data_tumor$metastasis
gsurv(FU, outcome, status) + ggtitle("Presence of metastatic disease after cystectomy \nand overall mortality")
# 5hmC and tumor progression
status <- data_tumor$progression
gsurv(FU, outcome, status) + ggtitle("Presence of tumor progression and \noverall mortality")
```

***

### Cancer-related mortality
```{r Clinicopath_DOD}
# Establishing outcome
outcome <- data_tumor$dod
# Patient's agedata_tumor$age_group
status <- data_tumor$age_group
gsurv(FU, outcome, status) + ggtitle("Patient's age and cancer-related mortality")
# Patient's sex
status <- data_tumor$sex
gsurv(FU, outcome, status) + ggtitle("Patient's sex and cancer-related mortality")
# Patient's race
status <- data_tumor$race
gsurv(FU, outcome, status) + ggtitle("Patient's race and cancer-related mortality") + theme(legend.position = c(.85, .9))
# pT stage at cystectomy
status <- data_tumor$mid
gsurv(FU, outcome, status) + ggtitle("Presence of muscle-invasive (>pT1) tumor and \ncancer-related mortality")
# Lymphovascular invasion
status <- data_tumor$lvi
gsurv(FU, outcome, status) + ggtitle("Presence of lymphovascular invasion and \ncancer-related mortality")
# Lymph nodes metastases
status <- data_tumor$ln.met
gsurv(FU, outcome, status) + ggtitle("Presence of lymph node metastases and \ncancer-related mortality")
# Early complications
status <- data_tumor$comp.early
gsurv(FU, outcome, status) + ggtitle("Presence of early complications and \ncancer-related mortality")
# Late complications
status <- data_tumor$comp.late
gsurv(FU, outcome, status) + ggtitle("Presence of late complications and \ncancer-related mortality")
# Metastatic disease after cystectomy
status <- data_tumor$metastasis
gsurv(FU, outcome, status) + ggtitle("Presence of metastatic disease after cystectomy \nand cancer-related mortality")
# 5hmC and tumor progression
status <- data_tumor$progression
gsurv(FU, outcome, status) + ggtitle("Presence of tumor progression and \ncancer-related mortality")
```

## 5hmC levels and survival: Approach #1
In this approch we will plot `Kaplan-Meier curves` for overall mortality and cancer-related mortality using the `median` H-score as the cutoff point for defining low/high 5hmC levels. We will compare survival curves using the `log-rank test`.

```{r Approach1}
# Splitting in 2 levels based on median score
data_tumor <- data_tumor %>%
  mutate(p50 = ifelse(t.5hmc.med > median(t.5hmc.med, na.rm = TRUE), "High", "Low")) %>%
  droplevels()
data_tumor$p50 <- factor(data_tumor$p50)
# 5hmC and overall mortality
outcome <- data_tumor$death
status <- data_tumor$p50
gsurv(FU, outcome, status) + ggtitle("5hmC levels and overall mortality")
# 5hmC and cancer-related mortality
outcome <- data_tumor$dod
status <- data_tumor$p50
gsurv(FU, outcome, status) + ggtitle("5hmC levels and cancer-related mortality")
```

***

## 5hmC levels and survival: Approach #2
In this approch we will plot `Kaplan-Meier curves` for overall mortality and cancer-related mortality using the upper tertile `Q3` H-score as the cutoff point for defining low/high 5hmC levels. We will compare survival curves using the `log-rank test`.

```{r Approach2}
# Splitting in 2 levels based on Q3 score
data_tumor <- data_tumor %>%
  mutate(q3 = ifelse(t.5hmc.med > quantile(t.5hmc.med, .66, na.rm = TRUE), "High", "Low")) %>%
  droplevels()
data_tumor$q3 <- factor(data_tumor$p50)
# 5hmC and overall mortality
outcome <- data_tumor$death
status <- data_tumor$q3
gsurv(FU, outcome, status) + ggtitle("5hmC levels and overall mortality")
# 5hmC and cancer-related mortality
outcome <- data_tumor$dod
status <- data_tumor$q3
gsurv(FU, outcome, status) + ggtitle("5hmC levels and cancer-related mortality")
```

***
