---
title: Expression and Regression Models
subtitle: 5mhC expression in bladder cancer
output: 
  html_document: 
    keep_md: yes
---

```{r Global, echo = FALSE, warning = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(knitr.table.format = "markdown")
library(plyr)
library(dplyr)
library(tidyr)
source("gplots.R")
```

```{r Data}
# Loading the dataset, libraries and sources for analysis
data <- read.delim("DatasetBladder.csv")
# Lumping pT levels at cystectomy considering muscle invasion
data <- data %>%
  mutate(mid = ifelse(pt.cystectomy == "Tis-Ta" | pt.cystectomy == "T1", "No", "Yes"))
data$mid <- factor(data$mid)
```

```{r Regression}
# Logistic regression tables
logistic.table <- function(outcome, predictors, varlabels){
  model <- list()
  OR.center <- list()
  OR.low <- list()
  OR.high <- list()
  OR.p <- list()
  for(i in 1:length(predictors)) {
    model[[i]] <- glm(outcome ~ predictors[[i]], family = binomial)
    OR.center[[i]] <-format(exp(coef(model[[i]]))[2], digits = 2, nsmall = 2)
    OR.low[[i]] <- format(exp(confint(model[[i]]))[2, 1], digits = 2, nsmall = 2)
    OR.high[[i]] <- format(exp(confint(model[[i]]))[2, 2], digits = 2, nsmall = 2)
    OR.p[[i]] <- format(summary(model[[i]])$coef[2, 4], digits = 2, width  = 6)
  }
  OR.center <- unlist(OR.center)
  OR.low <- unlist(OR.low)
  OR.high <- unlist(OR.high)
  OR.p <- unlist(OR.p)
  OR.table <- cbind("Variables" = varlabels, "OR" = OR.center,
    "Lower 95% CI" = OR.low, "Upper 95% CI" = OR.high, "P value" = OR.p)
  knitr::kable(OR.table, row.names = FALSE, align = c("l", "c", "c", "c", "c"))
}
# Cox regression tables
cox.table <- function(outcome, fu, predictors, varlabels){
  outcome <- as.numeric(outcome)
  survival.obj <- survival::Surv(fu, outcome)
  model <- list()
  HR.center <- list()
  HR.low <- list()
  HR.high <- list()
  HR.p <- list()
  for(i in 1:length(predictors)) {
    model[[i]] <- survival::coxph(survival.obj ~ predictors[[i]])
    HR.center[[i]] <-format(summary(model[[i]])$conf.int[1], digits = 2, nsmall = 2)
    HR.low[[i]] <- format(summary(model[[i]])$conf.int[3], digits = 2, nsmall = 2)
    HR.high[[i]] <- format(summary(model[[i]])$conf.int[4], digits = 2, nsmall = 2)
    HR.p[[i]] <- format(summary(model[[i]])$logtest[3], digits = 2, width  = 6)
  }
  HR.center <- unlist(HR.center)
  HR.low <- unlist(HR.low)
  HR.high <- unlist(HR.high)
  HR.p <- unlist(HR.p)
  HR.table <- cbind("Variables" = varlabels, "HR" = HR.center,
    "Lower 95% CI" = HR.low, "Upper 95% CI" = HR.high, "P value" = HR.p)
  knitr::kable(HR.table, row.names = FALSE, align = c("l", "c", "c", "c", "c"))
}
```


In this section we will build regression models for overall mortality and cancer-related mortality according to clinicopathologic features and 5hmC expression, categorizing 5hmC expression in low/high levels using as cutoff points: 1) the median; 2) the upper tertile. Only invasive cases will be consider, as observed at the TMA spot (i.e., `spot.dx == Invasive`). For age (numeric variable) we will use the median to define younger and older age groups.

```{r Invasive}
data_tumor <- data %>%
  filter(spot.dx == "Invasive") %>%
  mutate(age_group = ifelse(age > median(age), "Older", "Younger")) %>%
  droplevels()
data_tumor$age_group <- factor(data_tumor$age_group, ordered = TRUE)
# Splitting in 2 levels based on median score
data_tumor <- data_tumor %>%
  mutate(p50 = factor(ifelse(t.5hmc.med > median(t.5hmc.med, na.rm = TRUE), "High", "Low"))) %>%
  droplevels()
# Splitting in 2 levels based on Q3 score
data_tumor <- data_tumor %>%
  mutate(q3 = factor(ifelse(t.5hmc.med > quantile(t.5hmc.med, .66, na.rm = TRUE), "High", "Low"))) %>%
  droplevels()
```

We will begin building uncondicional binary logistic regression models to estimate the odds ratios. We then will build Cox's proportional hazards regression models to estimate the hazard ratios. We will consider time from cystectomy for the follow-up.

***

## Logistic regression: Overall mortality

```{r OR_death}
outcome <- data_tumor$death
predictors <- with(data_tumor, list(
  "Age group: Older vs. Younger" = age_group,
  "Sex" = sex,
  "Race" = race,
  "Neoadjuvant radiotherapy" = pre.rt,
  "Neoadjuvant chemotherapy" = pre.ct,
  "Intravesical chemotherapy" = intra.ct,
  "Adjuvant radiotherapy" = pos.rt,
  "Adjuvant chemotherapy" = pos.ct,
  "Early complications" = comp.early,
  "Late complications" = comp.late,
  "Lymph node metastasis" = ln.met,
  "Distant metastasis" = metastasis,
  "Lymphovascular invasion" = lvi,
  "Muscle-invasive disease" = mid,
  "High 5hmC expression (upper median)" = p50,
  "High 5hmC expression (upper tertile)" = q3
  ))
varlabels <- names(predictors)
logistic.table(outcome, predictors, varlabels)
```

***

## Logistic regression: Cancer-related death

```{r OR_dod}
outcome <- data_tumor$dod
logistic.table(outcome, predictors, varlabels)
```

***

## Cox regression: Overall mortality

```{r HR_death}
# Establishing follow-up
outcome <- data_tumor$death
fu <- data_tumor$fu.cystectomy
cox.table(outcome, fu, predictors, varlabels)
```

***

## Cox regression: Cancer-related mortality

```{r HR_dod}
outcome <- data_tumor$dod
cox.table(outcome, fu, predictors, varlabels)
```

